Physcraper: A Python package for continually updated phylogenetic trees using the Open Tree of Life

### 2021-01-07 sources/bmc_template_submission/bmc_article.tex
discussion and bib

Physcraper is the first tool for interoperability in phylogenetic analyses (???)
As a tool for automatization of phylogenetic reconstruction from molecular databases,
we believe it presents several advantages over existing phylogenetic pipelines designed to
make evolutionary sense of the vast amount of
public molecular data. Phylota \cite{sanderson2008phylota},
and PHLAWD \cite{smith2009mega} are the first ones and have been cited and used
the most for actual phylogenetic reconstruction and mining of databases. These are
strong tools that will probably still have a great potential in their recent updated versions
(phylotar \cite{bennett2018phylotar} and PyPHLAWD \cite{smith2019pyphlawd}).
They focus on generating full phylogenies \textit{de novo},
i.e., inferring phylogenetic relationships from a newly generated homology hypothesis,
as opposed to e.g., supertrees, that are generated by summarizing previous
phylogenetic estimates, which requires more manual adjustment later and difficults
comparison with previous phylogenetic information.

SUPERSMART \cite{antonelli2017toward}) and PUMPER \cite{izquierdo2014pumper}

The PUMPER pipeline \cite{izquierdo2014pumper} also uses the concept of updating
pre-existing alignments to incorporate public molecular data into phylogenies.
Unfortunately, installation was unsuccessful following instructions from the author,
and a comparison analysis between Physcraper and PUMPER is unfeasible at present.
%% **MTH: add some more details about what aspect of the PUMPER install failed**

Physcraper generates individual gene trees, failing to capture the complexity of
species' evolutionary history \cite{song2012resolving}. Yet, Physcraper facilitates
gathering alignments and gene trees for multiple loci from a group of interest,
that can be used to reconstruct species trees with ASTRAL \cite{mirarab2014astral},
BEAST2 \cite{bouckaert2019beast}, or SVD Quartets \cite{chifman2014quartet}).

Physcraper can potentially link phylogenies to data available in any of the
taxonomies integrated in the OpenTree taxonomy \cite{rees2017automated}, such as
geographical locations from the Global Biodiversity Information Facility, or
fossils from the Paleobiology Database.
%% a resource that is contributing to achieve the goal of "exploring and
%% analysing biodiversity at an accelerated pace, and returning systematics into
%% the mainstream of science" \cite{wilson2003encyclopedia}.
The Physcraper workflow can be used to rapidly (in a matter of hours)
address challenges overarching both fields of ecology and evolution, such as
phylogenetically placing newly discovered species \cite{webb2010biodiversity},
systematizing molecular (and other) databases, i.e., curating taxonomic assignations
\cite{san2010molecular},
and generating custom trees for ecological \cite{helmus2012phylogenetic} and
evolutionary downstream analyses \cite{stoltzfus2013phylotastic}.

>>>>

To our knowledge, Physcraper is the first tool establishing an interoperability
framework for phylogenetic analyses (should we say this???).
We believe this is a key step to successfully establish an open, reproducible workflow for phylogenetics,
facilitating phylogenetic knowledge for ecologists and other non-specialists, effectively
democratizing phylogenetic studies.

As a tool for automatizing phylogenetic reconstruction from molecular databases,
we believe it presents several advantages over existing phylogenetic pipelines designed to
make evolutionary sense of the vast amount of
public molecular data available. Phylota \cite{sanderson2008phylota},
and PHLAWD \cite{smith2009mega} were the first of the kind and have been cited and used
abundantly for phylogenetic reconstruction and mining of molecular databases. While these are
strong tools that will probably still be widely used with great potential in their
more recent updated versions (phylotar \cite{bennett2018phylotar} and PyPHLAWD \cite{smith2019pyphlawd}),
they do not take into account prior work and knowledge on phylogenetic relationships
and alignment construction, which makes them prone to error (requiring more
manual downstream processing) and difficults comparison with previous phylogenetic
knowledge.

SUPERSMART \cite{antonelli2017toward}) and PUMPER \cite{izquierdo2014pumper}
share conceptual strengths, but present many challenges for installation and user
experience, which have resulted in a low level of adoption of these tools
in the community, and have been applied in very few phylogenetic analyses since their
publication (put number of citations as of today or to harsh??).

Phylogenerator \cite{pearse2013phylogenerator} gathers genetic data by relying only on
genetic marker names and descriptions in the GenBank database, which can be prone
to error due to misslabelling. Any form of sequence similarity analysis, such as BLAST
\cite{camacho2009blast} greatly improves correct homologous identification and should
be preferred for obtaining high-quality phylogenetic hypotheses.

While Physcraper generates individual gene trees, which generally fail to capture the complexity of
species' evolutionary history \cite{song2012resolving}, it facilitates
gathering alignments and gene trees for multiple loci from a group of interest,
that can be used to reconstruct species trees with ASTRAL \cite{mirarab2014astral},
BEAST2 \cite{bouckaert2019beast}, or SVD Quartets \cite{chifman2014quartet}).

Physcraper has the unique advantage to link phylogenies to data available in any of the
taxonomies integrated in the OpenTree taxonomy \cite{rees2017automated}, such as
geographical locations from the Global Biodiversity Information Facility, or
fossils from the Paleobiology Database.
%% a resource that is contributing to achieve the goal of "exploring and
%% analysing biodiversity at an accelerated pace, and returning systematics into
%% the mainstream of science" \cite{wilson2003encyclopedia}.
The Physcraper workflow can be used to rapidly (in a matter of hours)
address challenges overarching both fields of ecology and evolution, such as
phylogenetically placing newly discovered species \cite{webb2010biodiversity},
systematizing molecular (and other) databases, i.e., curating taxonomic assignations
\cite{san2010molecular},
and generating custom trees for ecological \cite{helmus2012phylogenetic} and
evolutionary downstream analyses \cite{stoltzfus2013phylotastic}.
